The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-01-04 pkg/partykit/inst/guideliketest/draft/draft.Rnw
small corrections, mainly in sections 2 and 3; structured discussion
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2401 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9f7d5a8831aafb19fea0a98bde255980387338dd

<<preliminaries, echo=FALSE, results=hide>>=
# setwd("~/svn/partykit/pkg/partykit/inst/guideliketest/draft/")
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
library("partykit")
library("Formula")
library("parallel")
library("lattice")
library("scales")
# setwd("~/svn/partykit/pkg/partykit/inst/guideliketest/draft/")
source("../guidelike.R")
source("../tests.R")
#load("../sim/simres20180422_1step.rda")
load("../sim/simres20181223_stump.rda")
sim_stump <- simres$res
#load("../sim/simres20180422_2step.rda")
#sim2step <- simres
rm(simres)
load("../sim/simres20181223_tree.rda")
sim_tree <- simres$res
rm(simres)

@@

In this paper the focus will be put on this first step, i.e. the task of selecting 
the right/best split variable. In particular, the approach of the GUIDE algorithm is
compared to the one of the CTree algorithm and the MOB algorithm. For this purpose, a 
unifying framework for unbiased model based tree algorithms is presented such that each 
of the three algorithms can be obtained by a specific combination of the available building 
blocks. In that way, the effects of each of these characteristic features of the GUIDE,
CTree and MOB algorithm can be investigated separately by varying the switches in
the tool box provided by this general framework. 
Depending on the setup different effects can be observed for the investigated characteristics.
However, none of the results contradicts the conclusion that in general including all available 
model scores is at least as efficient but mostly clearly advantageous over considering only 
residuals. Moreover, the effects of binarizing residuals or scores at $0$ or categorizing
the values of possible split variables are investigated and the impact of applying a conditional
or an unconditional approximation of the null distribution is analyzed.

In the first part of Section~\ref{sec:urp} a basic scheme on how to build a tree is given. 
Then, after explaining the investigated testing strategies more in detail in 
Sections~\ref{sec:ctree}, \ref{sec:mob} and \ref{sec:guide}, the unifying framework for model 
based unbiased recursive partitioning is presented in Section~\ref{sec:unify}.
The setting for the simulation study is introduced in Section~\ref{sec:simulation}
and the results are discussed in Section~\ref{sec:results}.

@@

\subsection{Generic Algorithm}
The basic idea of building a regression tree model is to partition the data into smaller and 
more homogeneous subgroups based on (a set of) covariates. Various different tree algorithms 
have been developed, all following the same general structure:
\begin{enumerate}
\item For the whole underlying data set in the current node some form of discrepancy 
measure is calculated. 
%This can for example be the sum of deviations from a typical or average value of the 
%data set or it can be based on a fitted model.
\item Among all available covariates one is selected as split variable and a value 
within the range/measurable space of this variable is chosen as split point such that 
after splitting the data the discrepancy measure is minimal within the resulting subgroups.
\item Steps 1 and 2 are repeated within each new subgroup until either a stopping criterion
is reached or the subgroup consists of only one or all equal observations. (In the second
case the resulting tree is of maximal size and can be reduced by (post-)pruning.)
\end{enumerate}

The employed discrepancy measure can for example be the sum of deviations from a typical 
or average value, but it can also be constructed based on a statistical model. For instance,
different types of residuals or their signs can be applied (see e.g., \cite{Loh:2002}) as well as
rank sums or logrank scores as in \cite{Hothorn+Hornik+Wiel:2006}.
(In the simulation study of this paper trees with linear models fitted in each node are 
investigated with the residual sum of squares (RSS) employed as discrepancy measure. However, 
the discussed methodology and the presented results also apply for a wide range of other 
model types.)

As explained in Section~\ref{sec:introduction} the considered tree algorithms CTree, MOB, and 
GUIDE first select the split variable and then the split point in a separate step. 
For the first step of selecting a split variable they all apply statistical (independence) tests
following the same general strategy: 
Based on the discrepancy measure an $n \times k$ - matrix $D$ is obtained giving information
on the goodness of fit of the model for a data set of $n \in \mathbb{N}$ observations in the 
current node and $k \in \mathbb{N}$.
Using this matrix, the algorithm-specific test statistic is calculated for each possible split 
variable $Z_j$, $j \in \{1,\ldots,m\}$ for $m \in \mathbb{N}$, evaluating the dependency
of $Z_j$ on $D$. As the variables are usually not all measured at the same scale the resulting
values of the test statistics can not be compared directly in an unbiased way. Therefore, it is
necessary to calculate the corresponding $p$-values in order to allow for an unbiased
variable selection on this standardized/unified scale. Then the variable corresponding to the smallest
$p$-value is selected as split variable as it has shown to be the variable with the strongest
influence on the goodness of fit of the model.
% Hinweis: pruning-Strategien
%Once the split variable is chosen the corresponding split point is then selected such that it 
%minimizes the discrepancy measure.

@@

The CTree algorithm is based on the idea of providing non-parametric regression tree models in 
a conditional inference framework by applying permutation tests. 
To select a split variable it is tested whether there is any association between the transformed 
response $h(Y)$ and $g(Z)$ for a so-called influence function $g$ and each possible split variable $Z$.
For different types of response variables different functions $h$ which depend on $Y$ in a 
permutation-symmetric way and might also be multidimensional can be selected.
For a numeric response the identity function $h(Y) = Y$ is a common choice, while a categorical
response can be mapped to a unity vector indicating the corresponding category by an indicator
function $h(Y) = (0,\ldots,1,\ldots,0)^{\top}$. Alternatively, the function $h$ can return a location
and scale parameter of $Y$, e.g. $h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
If a parametric model is fit to the values of the response variable~$Y$ and regressor variable(s)~$X$
the corresponding model scores can be obtained by setting $h(Y) = s(Y,X,\hat{\theta})$
where $\hat{\theta}$ is the estimated model parameter and $s$ is the score function:
$$
s(Y, X, \hat{\theta}) = \frac{\partial \psi(Y, X, \hat{\theta})}{\partial \theta}
$$
with $\psi$ being the objective function which is minimized to estimate the model parameter
$\theta$. Hence, for a data set of $n$ observations containing the response vector 
$y = (y_1,\ldots,y_n)^\top$ and regressor(s) $x = (x_1,\ldots,x_n)^\top$ and a $k$-dimensional 
parameter vector $\theta$ the score function $s$ returns a $n \times k$ - matrix.
In that way a score value indicating how well the model fits the data is given separately
for each tuple of observation and parameter element $(y_i, x_i, \beta_l)$ with $i \in \{1,\dots,n\}$
and $l \in \{1,\ldots,k\}$. 
(To estimate $\theta$ many different M-type estimators can be applied including maximum likelihood 
or ordinary least squares. Therefore, a model-based discrepancy measure considered for the tree 
algorithm can be employed as objective function.)

Similarly, different types of functions can be chosen for the influence function $g$ depending
on a possible split variable $Z$. A simple choice in case of $Z$ being a numeric variable is again
the identity function $g(Z) = Z$. For categorical variables $g$ can also map its values to the 
corresponding unity vectors by $g(Z) = (0,\ldots,1,\ldots,0)$.

@@

\subsection{MOB}\label{sec:mob}
While CTree offers to apply model-based discrepancy measures as one possible choice, the MOB algorithm 
was developed with the aim of embedding parametric models into the framework of regression trees
and is hence purely intended for a model-based approach. The testing strategy for selecting a split
variable is based on a fitted model, i.e. on an estimated parameter (vector) $\hat{\theta}$ for each
node of the tree, and on the application of the Central Limit Theorem. However, the fitted parametric 
models can be of a very basic and simple form such as an average value. The algorithm only requires the 
corresponding model scores $s(Y,X,\hat{\theta})$ in order to test for any instability in the model 
parameters over each particular ordering induced by each of the possible partitioning variables $Z$.
If a variable $Z$ has no specific influence on the parameters the scores would always fluctuate randomly
around their mean 0. Therefore, these deviations are investigated by measuring them using an 
empirical fluctuation process. Applying a scalar function which captures the fluctuation in this process 
yields the test statistic of this so called generalized M-fluctuation test. To obtain the distribution of
the resulting test statistic asymptotic theory, in particular the Central Limit Theorem is applied.
In this parametric unconditional framework $p$-values can be calculated and the covariate
with the smallest $p$-value is selected as split variable. 
For $Z$ being a numeric variable a maximally selected split point search is employed while for a 
categorical variable all possible splits are evaluated to find the best one.
Hence, the number of possible breakpoints is not fixed unlike in the CTree algorithm which can make 
this operation computationally expensive.

\subsection{GUIDE}\label{sec:guide}
Similar to the MOB algorithm, parametric models are fit to the nodes of a tree in the GUIDE algorithm.
For the statistical test applied to select a split variable in each node it requires some type of 
residuals $r(Y,X,\hat{\theta})$ from the models such as for example 
$r(Y,X,\hat{\theta}) = Y - X^{\top}\hat{\theta}$. 
Before applying a statistical test two additional steps are performed:
The residuals are binarized, i.e. depending on whether the sign of a residual is 
positive or not it is replaced by $1$ or $2$  % FIX ME: +1 and -1 in the implementation? 
and for each possible split variable the observed values are categorized into 4 groups 
separated by the corresponding 1st, 2nd and 3rd quartile.
Then, a $\chi^2$-test of independence is performed for the binarized residuals and
each categorized split variable.
After choosing the split variable showing the highest dependency by yielding the lowest $p$-value,
the split point minimizing the discrepancy measure is selected. The possible number of breakpoints
is determined by the categorization of the possible split variables such that the quartiles are
the only three possible split points.

@@

\textit{Note:}
While GUIDE originally constructs large trees and applies post-pruning with the cross-validation 
method of CART, CTree and MOB employ pre-pruning by setting a significance level for the 
independence tests such that no split is performed if the lowest $p$-value is not smaller than 
the significance level. The impact of the choice of pruning strategy is further discussed in 
Section~\ref{sec:discussion}.

\section{Unifying Framework}\label{sec:unify}
As all three algorithms follow the same basic structure their approaches can be presented together
in a unifying framework for a model based tree algorithm from which each specific algorithm can be 
obtained by switching on or off certain features. By applying such a general framework each of the 
characteristic features of the three algorithms and its impact can be investigated in detail. 
These switches are explained in this Section and their setting for the algorithms
CTree, MOB, and GUIDE are listed in Table~\ref{tab:switches}.

\subsection{Switches}
\begin{itemize}
\item  Transformations of Y:\\
The selection of a split variable is based on a statistical test investigating the dependency of the
response $Y$ on each possible split variables $Z$. However, before evaluating such a test a model-based
transformation of $Y$ is applied to obtain a discrepancy measure. This can be residuals 
$r(Y,X,\hat{\theta})$ as in the GUIDE algorithm or scores $s(Y,X,\hat{\theta})$ as in the MOB algorithm
or, more generally, a function $h(Y,X)$ as in the CTree algorithm including both of the foregoing cases.
Since the first column of the score matrix $S = s(Y,X,\hat{\theta})$ is the vector of residuals the 
GUIDE algorithm selects only this first column while the MOB algorithm uses all available score values,
hence the full model matrix $S$.
%providing information on the goodness of fit for each observation and model parameter.
(By including all available scores the goodness of fit of the model is investigated regarding 
each parameter rather than just regarding the mean/location parameter. In that way changes in 
parameters other than the mean/location parameter should be detected more easily.)

@@

\item Categorization of covariates:\\
Similarly, the covariates can also be categorized. For this purpose, the corresponding 1st, 2nd 
and 3rd quartile are used to bin each possible split variable into four categories. In that way
a $\chi^2$-test can be applied and the quartiles are the only possible split points while otherwise maximally selected splits over $Z$ are considered.

\end{itemize}
%\subsubsection{Conditional vs unconditional}

@@

\subsection{Linear model tree}
As an example a linear model tree is considered as this allows 
for an easy and well structured description of the investigated subjects/issues. 
For that reason, this model is also considered for the simulation study in Section~\ref{sec:simulation}.
However, the conclusions drawn for this example also apply for a wide range of other model
types. The models fitted in the nodes of the considered linear tree model are of the form
$$
Y = \beta_0 + \beta_1 \cdot X + \epsilon
$$
for a response variable $Y$, a regressor variable $X$ and an error $\epsilon$. 
As for each node the type of the fitted model is the same no node-specific features are considered
and the observations of the current (arbitrarily chosen) node are denoted by 
$\{(y_i,x_i)\}_{i=1,\ldots,n}$ for $n \in \mathbb{N}$.

A very basic form of a discrepency measure can be obtained by simply calculating deviations of the 
response values from the mean value within the node, for example by $\sum_{i=1}^n (\bar{y} - y_i)^2$.
However, this includes only the response variable $Y$ but not the fitted model.
For a model-based approach residual can be considered such as in the residual sum of squares (RSS)
$$
r(y,x,\boldsymbol{\beta}) = RSS(y, x, \boldsymbol{\beta}) = 
\sum_{i=1}^n (y_i - \beta_0 - \beta_1 \cdot x_i)^2.
$$ 
which is also the objective function that is minimized for the estimation of the coefficient vector 
$\boldsymbol{\beta} = (\beta_0, \beta_1)^\top$, 
i.e. to obtain $\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
The corresponding scores are defined by
$$
s(y_i, x_i, \hat{\boldsymbol{\beta}}) = 
\frac{\partial \textit{RSS}}{\partial \boldsymbol{\beta}}(y_i, x_i, \hat{\boldsymbol{\beta}})
$$
such that the score matrix $S$ is in this case a $n \times 2$ - matrix as $\boldsymbol{\beta}$ is 
two dimensional. In that way a score value indicating how well the model fits the data is given separately
for each tuple of observation and coefficient $(y_i,x_i,\beta_j)$ with $i \in \{1,\dots,n\}$
and $j \in \{1,2\}$.

@@

The location parameter $\mu$ of the normally distributed response variable $Y$ depends
linearly on the regressor variable $X$ which is a uniformly distributed variable taking 
values between $-1$ and $1$.
% can either be a binary variable (taking either $0$ or $1$) or a uniformly distributed 
% variable (taking values between $-1$ and $1$).
Moreover, the intercept $\beta_0$ and the slope parameter $\beta_1$ of this linear equation
(can) depend on the true split variable $Z_1$. 
%in the following way:
%\begin{center}
%$\mu(X, Z_1) = \beta_0(Z_1) + \beta_1(Z_1) \cdot X$.
%\end{center}
%
In particular, three different cases are considered for the coefficients $\beta_0$ 
and $\beta_1$:
\begin{enumerate}
\item The intercept $\beta_0$ varies depending on $Z_1$ while $\beta_1$ is fixed (at $1$).
\item The slope coefficient $\beta_1$ varies depending on $Z_1$ while $\beta_0$ is fixed 
(at $0$).
\item Both coefficients $\beta_0$ and $\beta_1$ vary depending on $Z_1$.
\end{enumerate}

@@

\begin{figure}%[h!]
\setkeys{Gin}{width=1\linewidth}
\minipage{0.45\textwidth}
<<dgp_stump1, fig=TRUE, echo=FALSE, width=4, height=4>>=
#par(mar=c(3,3,0,0), oma = c(3,3,0,0))
plot(p, 
     tp_args = list(FUN = identity, width = 15, fill = paltrees[c(1, 3)]), 
     ip_args = list(fill = paltrees[c(2, 3)]),
     drop_terminal = TRUE, tnex = 1, edge_panel = ep)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_stump2, fig=TRUE, echo=FALSE, width=4, height=4>>=
par(mar = c(4,4,2,1))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", 
     main = expression("varying"~beta[0]),
     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
abline(coef = c(-1,1), col = "blue", lwd = 2)
abline(coef = c(1,1), col = "red", lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha("red", 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
@
\endminipage

@@

Moreover, all of the considered statistical tests are applied as significance tests answering 
two questions at once: (1) Should a split be performed at all? (2) If so, in which variable?
For the first question the $p$-values regarding all available covariates are compared 
to a predefined level of significance $\alpha$. Only if the smallest $p$-value is smaller
than $\alpha$ a split is performed and the covariate corresponding to this $p$-value is 
selected. Including this criterion of significance in the building process of a tree is often
referred to as \textit{pre-pruning}, however, this is not always applied in practice as 
significance is often not considered. As mentioned before, the GUIDE algorithm usually builds
large trees which are then pruned afterwards. However, for the purpose of comparing the
testing strategies rather than the whole tree algorithms in this simulation study a 
significance level is considered for all statistical tests, including the $\xi^2$-test in 
the GUIDE algorithm such that all algorithms apply pre-pruning rather than post-pruning.

To compare the performance of the evaluated testing strategies for each step the proportion
of data sets for which the true splitting variable $Z_1$ is detected and also selected 
(for a given significance level $\alpha = 0.05$) is considered for the first scenario 
(stump).
For the second scenario (tree with 2 splits) the adjusted Rand index (ARI) is calculated 
as a measure of similarity between the true tree structure and the model tree.

@@

\subsection{Tree}
\label{sec:tree}
To compare the performance of testing strategies in a more complex scenario they have also
been evaluated on data sets based on a tree structure with two different split variables 
($Z_1$ and $Z_2$), both with the same split point $\xi$ as illustrated in 
Figure~\ref{fig:dgp_tree}.

>>>>

<<preliminaries, echo=FALSE, results=hide>>=
# setwd("~/svn/partykit/pkg/partykit/inst/guideliketest/draft/")
options(prompt = "R> ", continue = "+  ", useFancyQuotes = FALSE, width = 70)
library("partykit")
library("Formula")
library("parallel")
library("lattice")
library("scales")
# setwd("~/svn/partykit/pkg/partykit/inst/guideliketest/draft/")
source("../guidelike.R")
source("../tests_guideprune.R")
#load("../sim/simres20180422_1step.rda")
load("../sim/simres20181223_stump.rda")
sim_stump <- simres$res
#load("../sim/simres20180422_2step.rda")
#sim2step <- simres
rm(simres)
load("../sim/simres20181223_tree.rda")
sim_tree <- simres$res
rm(simres)

@@

In this paper the focus will be put on this first step, i.e. the task of selecting 
the best split variable. In particular, the approach of the GUIDE algorithm is
compared to the one of the CTree algorithm and the MOB algorithm. For this purpose, a 
unifying framework for unbiased model-based tree algorithms is presented such that each 
of the three algorithms can be obtained by a specific combination of the available building 
blocks. In that way, the effects of each of these characteristic features of the GUIDE,
CTree and MOB algorithm can be investigated separately by varying the switches in
the tool box provided by this general framework. 
Depending on the setup different effects can be observed for the investigated characteristics.
However, none of the results contradicts the conclusion that in general using all available 
model scores is at least as efficient but mostly clearly advantageous over considering only 
residuals. Moreover, the effects of binarizing residuals or scores at $0$ or categorizing
the values of possible split variables are investigated and the impact of applying a conditional
or an unconditional approximation of the null distribution is analyzed.

In the first part of Section~\ref{sec:urp} a basic scheme on how to build a tree is given. 
Then, after explaining the investigated testing strategies more in detail in 
Sections~\ref{sec:ctree}, \ref{sec:mob} and \ref{sec:guide}, the unifying framework for 
model-based unbiased recursive partitioning is presented in Section~\ref{sec:unify}.
The setting for the simulation study is introduced in Section~\ref{sec:simulation}
and the results are discussed in Section~\ref{sec:results}.

@@

\subsection{Generic Algorithm}
The basic idea of building a regression tree model is to partition the data into smaller and 
more homogeneous subgroups based on (a set of) covariates. Various different tree algorithms 
have been developed, all following the same general structure starting at the root of the tree
which is simply the first node containing all available data:
\begin{enumerate}
\item For the whole underlying data set in the current node some form of discrepancy 
measure is calculated. 
%This can for example be the sum of deviations from a typical or average value of the 
%data set or it can be based on a fitted model.
\item Among all available covariates one is selected as split variable and a value 
within the range/measurable space of this variable is chosen as split point such that 
after splitting the data into subgroups the discrepancy measure is minimal within the 
resulting nodes.
\item Steps 1 and 2 are repeated within each new node until either a stopping criterion
is reached or the data in the node consists of only one or all equal observations. 
\end{enumerate}

The employed discrepancy measure can for example be the sum of deviations from a typical 
or average value, but it can also be constructed based on a statistical model. For instance,
different types of residuals or their signs can be applied (see e.g., \cite{Loh:2002}) as well as
rank sums or logrank scores as in \cite{Hothorn+Hornik+Wiel:2006}.
(In the simulation study of this paper trees with linear models fitted in each node are 
investigated with the residual sum of squares (RSS) employed as discrepancy measure.)

As explained in Section~\ref{sec:introduction} the considered tree algorithms CTree, MOB, and 
GUIDE first select the split variable and then the split point in a separate step. 
For the first step of selecting a split variable they all apply statistical tests
following the same general strategy: 
Based on the discrepancy measure an $n \times k$-matrix $D$ is obtained giving information
on the goodness of fit of the model for a data set of $n \in \mathbb{N}$ observations in the 
current node and $k \in \mathbb{N}$.
Using this matrix, the algorithm-specific test statistic is calculated for each possible split 
variable $Z_j$, $j \in \{1,\ldots,m\}$ for $m \in \mathbb{N}$, evaluating the dependency
of $Z_j$ on $D$. As the variables are usually not all measured at the same scale the resulting
values of the test statistics can not be compared directly in an unbiased way. Therefore, it is
necessary to calculate the corresponding $p$-values in order to allow for an unbiased
variable selection on this standardized/unified scale. Then the variable corresponding to the smallest
$p$-value is selected as split variable as it has shown to be the variable with the strongest
influence on the goodness of fit of the model.

@@

The CTree algorithm is based on the idea of providing non-parametric regression tree models in 
a conditional inference framework by applying permutation tests. 
To select a split variable it is tested whether there is any association between the transformed 
response $h(Y)$ and $g(Z)$ for a so-called influence function $g$ and each possible split variable $Z$.
For different types of response variables different functions $h$ which depend on $Y$ in a 
permutation-symmetric way and might also be multidimensional can be selected.
For a numeric response the identity function $h(Y) = Y$ is a common choice, while a categorical
response can be mapped to a unity vector by an indicator function $h(Y) = (0,\ldots,1,\ldots,0)^{\top}$.
Alternatively, the function $h$ can return a location and scale parameter of $Y$, 
e.g. $h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
If a parametric model is fit to the values of the response variable~$Y$ and regressor variable(s)~$X$
the function $h$ can be a model-based transformation returning the corresponding model scores by 
setting $h(Y) = s(Y,X,\hat{\theta})$
where $\hat{\theta}$ is the estimated model parameter and $s$ is the score function
$$
s(Y, X, \hat{\theta}) = \frac{\partial \psi(Y, X, \hat{\theta})}{\partial \theta}
$$
with $\psi$ being the objective function which is minimized to estimate the model parameter
$\theta$. Hence, for a data set of $n$ observations containing the response vector 
$\boldsymbol{y} = (y_1,\ldots,y_n)^\top$ and 
regressor(s) $\boldsymbol{x} = (x_1,\ldots,x_n)^\top$ and 
a $k$-dimensional parameter vector $\boldsymbol{\theta} = (\theta_1,\ldots,\theta_k)^\top$ the 
score function $s$ returns a $n \times k$-matrix.
In that way a score value indicating how well the model fits the data is given separately
for each tuple of observation and parameter element $(y_i, x_i, \theta_l)$ with $i \in \{1,\dots,n\}$
and $l \in \{1,\ldots,k\}$. 
To estimate $\boldsymbol{\theta}$ many different M-type estimators can be applied including 
maximum likelihood or ordinary least squares. Therefore, these objective functions
are also possible options to obtain a model-based discrepancy measure in the tree algorithm.

Similarly, different types of functions can be chosen for the influence function $g$ depending
on a possible split variable $Z$. A simple choice in case of $Z$ being a numeric variable is again
the identity function $g(Z) = Z$. For categorical variables $g$ can also map its values to the 
corresponding unity vectors by an indicator function $g(Z) = (0,\ldots,1,\ldots,0)$.

@@

\subsection{MOB}\label{sec:mob}
While CTree offers to apply model-based discrepancy measures as one possible choice, the MOB algorithm 
was developed with the aim of embedding parametric models into the framework of regression trees
and is hence purely intended for a model-based approach. The testing strategy for selecting a split
variable is based on a fitted model, i.e. on an estimated parameter (vector) $\hat{\theta}$ for each
node of the tree, and on the application of the central limit theorem. However, the fitted parametric 
models can be of a very basic and simple form such as an average value. The algorithm only requires the 
corresponding model scores $s(Y,X,\hat{\theta})$ in order to test for any instability in the model 
parameters over each particular ordering induced by each of the possible partitioning variables $Z$.
If a variable $Z$ has no specific influence on the parameters the scores would always fluctuate randomly
around their mean 0. Therefore, these deviations are investigated by measuring them using an 
empirical fluctuation process. Applying a scalar function which captures the fluctuation in this process 
yields the test statistic of this so called generalized M-fluctuation test. To obtain the distribution of
the resulting test statistic asymptotic theory, in particular the central limit theorem is applied.
In this parametric unconditional framework $p$-values can be calculated and the covariate
with the smallest $p$-value is selected as split variable. 
For a categorical variable all possible splits are evaluated to find the best one while for $Z$ being 
a numeric variable a maximally selected split point search is employed. Hence, the number 
of possible breakpoints is not fixed unlike in the CTree algorithm which can make this operation 
computationally expensive.

\subsection{GUIDE}\label{sec:guide}
Similar to the MOB algorithm, parametric models are fit to the nodes of a tree in the GUIDE algorithm.
For the statistical test applied to select a split variable in each node some type of 
residuals from the models is required such as for example 
$r(Y,X,\hat{\theta}) = Y - X^{\top}\hat{\theta}$. 
Before applying a statistical test two additional steps are performed:
The residuals are binarized, i.e. depending on whether the sign of a residual is 
positive or not it is replaced by $1$ or $2$  % FIX ME: +1 and -1 in the implementation? 
and for each possible split variable the observed values are categorized into 4 groups 
separated by the corresponding 1st, 2nd and 3rd quartile.
Then, a $\chi^2$-test of independence is performed for the binarized residuals and
each categorized split variable.
After choosing the split variable showing the highest dependency by yielding the lowest $p$-value,
the split point minimizing the discrepancy measure is selected. The possible number of breakpoints
is determined by the categorization of the possible split variables such that the quartiles are
the only three possible split points.

@@

\textit{Note:}
While GUIDE originally constructs large trees and applies post-pruning with the cross-validation 
method of CART, CTree and MOB employ pre-pruning by setting a significance level for the 
independence tests such that no split is performed if the lowest $p$-value is not smaller than 
the significance level. As explained in Section~\ref{sec:evaluation}, within the simulation study 
in Section~\ref{sec:simulation} all three algorithms are applied using pre-pruning in order to put 
the focus on the power of the algorithms in terms of selecting the best split variable. 
However, the impact of the choice of pruning strategy is further discussed in Section~\ref{sec:discussion}.

\section{Unifying Framework}\label{sec:unify}
Since all three algorithms follow the same basic structure their approaches can be presented together
in a unifying framework for a model-based tree algorithm from which each specific algorithm can be 
obtained by switching on or off certain features. By applying such a general framework each of the 
characteristic features of the three algorithms and its impact can be investigated in detail. 
These switches are explained in this Section and their setting for the algorithms
CTree, MOB, and GUIDE are listed in Table~\ref{tab:switches}.

\subsection{Switches}
As explained before, the selection of a split variable is based on a statistical test 
investigating the dependency of the response $Y$ on each possible split variables $Z$.
The three following switches all refer to the way how the given data is prepared/transformed
before employing such a statistical test.
\begin{itemize}
\item  Transformations of Y:\\
A model-based transformation of $Y$ is applied to obtain a discrepancy measure giving 
information about the goodness of fit of the model. This can be residuals $r(Y,X,\hat{\theta})$
as in the GUIDE algorithm or scores $s(Y,X,\hat{\theta})$ as in the MOB algorithm or, 
more generally, a function $h(Y,X)$ as in the CTree algorithm including both of the foregoing cases.
Since the first column of the score matrix $S = s(Y,X,\hat{\theta})$ is the vector of residuals the 
GUIDE algorithm selects only this first column while the MOB algorithm uses all available score values,
hence the full model score matrix $S$.
%providing information on the goodness of fit for each observation and model parameter.
(By including all available scores the goodness of fit of the model is investigated regarding 
each parameter rather than just regarding the mean/location parameter. In that way changes in 
parameters other than the mean/location parameter should be detected more easily.)

@@

\item Categorization of covariates:\\
Similarly, the covariates can also be categorized. For this purpose, the corresponding 1st, 2nd 
and 3rd quartile are used to bin each possible split variable into four categories. In that way,
depending on whether the scores are binarized or not, a $\chi^2$-test or 1-way ANOVA can be applied 
where the quartiles are the only possible split points while otherwise maximally selected splits 
over $Z$ are considered.

\end{itemize}

\textit{Note:}
Apart from these three switches, the algorithms differ in their approximation of the underlying
null distribution of the test statistic. While the permutation tests in the CTree algorithm
work in a conditional framework the MOB and GUIDE algorithms both apply an unconditional approach.

@@

\subsection{Linear model tree}
As an example a linear model tree is considered as this allows 
for an easy and well structured description of the investigated subjects/issues. 
For that reason, this model is also considered for the simulation study in Section~\ref{sec:simulation}.
However, the conclusions drawn from this example also apply for a wide range of other model
types. The models fitted in the nodes of the considered linear tree model are of the form
$$
Y = \beta_0 + \beta_1 \cdot X + \epsilon
$$
for a response variable $Y$, a regressor variable $X$ and an error $\epsilon$. 
As for each node the type of the fitted model is the same no node-specific features are considered
and the observations of the current (arbitrarily chosen) node are denoted by 
$\{(y_i,x_i)\}_{i=1,\ldots,n}$ for $n \in \mathbb{N}$.

A very basic form of a discrepancy measure can be obtained by simply calculating deviations of the 
response values from the mean value within the node, for example by $\sum_{i=1}^n (\bar{y} - y_i)^2$.
However, this includes only the response variable $Y$ but not the fitted model.
For a model-based approach residuals can be considered such as in the residual sum of squares (RSS)
$$
r(y,x,\boldsymbol{\beta}) = RSS(y, x, \boldsymbol{\beta}) = 
\sum_{i=1}^n (y_i - \beta_0 - \beta_1 \cdot x_i)^2
$$ 
which is also the objective function that is minimized for the estimation of the coefficient vector 
$\boldsymbol{\beta} = (\beta_0, \beta_1)^\top$, 
i.e. to obtain $\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
The corresponding scores are defined by
$$
s(y_i, x_i, \hat{\boldsymbol{\beta}}) = 
\frac{\partial \textit{RSS}}{\partial \boldsymbol{\beta}}(y_i, x_i, \hat{\boldsymbol{\beta}})
$$
such that the score matrix $S$ is in this case a $n \times 2$-matrix as $\boldsymbol{\beta}$ is 
two dimensional. In that way a score value indicating how well the model fits the data is given separately
for each tuple of observation and coefficient $(y_i,x_i,\beta_j)$ with $i \in \{1,\dots,n\}$
and $j \in \{1,2\}$.

@@

The location parameter $\mu$ of the normally distributed response variable $Y$ depends
linearly on the regressor variable $X$.
% which is a uniformly distributed variable taking values between $-1$ and $1$.
% can either be a binary variable (taking either $0$ or $1$) or a uniformly distributed 
% variable (taking values between $-1$ and $1$).
Moreover, the intercept $\beta_0$ and the slope parameter $\beta_1$ of this linear equation
(can) depend on the true split variable $Z_1$. 
%in the following way:
%\begin{center}
%$\mu(X, Z_1) = \beta_0(Z_1) + \beta_1(Z_1) \cdot X$.
%\end{center}
%
In particular, three different cases are considered for the coefficients $\beta_0$ 
and $\beta_1$:
\begin{enumerate}
\item The intercept $\beta_0$ varies depending on $Z_1$ while $\beta_1$ is fixed (at $1$).
\item The slope coefficient $\beta_1$ varies depending on $Z_1$ while $\beta_0$ is fixed 
(at $0$).
\item Both coefficients $\beta_0$ and $\beta_1$ vary depending on $Z_1$.
\end{enumerate}

@@

\begin{figure}%[t]
\setkeys{Gin}{width=1\linewidth}
\minipage{0.45\textwidth}
<<dgp_stump1, fig=TRUE, echo=FALSE, width=4, height=4>>=
#par(mar=c(3,3,0,0), oma = c(3,3,0,0))
plot(p, 
     tp_args = list(FUN = identity, width = 15, fill = paltrees[c(1, 3)]), 
     ip_args = list(fill = paltrees[c(2, 3)]),
     drop_terminal = TRUE, tnex = 1, edge_panel = ep)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_stump2, fig=TRUE, echo=FALSE, width=4, height=4>>=
par(mar = c(4,4,2,1))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", 
     main = expression("varying"~beta[0]),
     pch = 19, col = alpha("blue", 0.4), cex = 0.7)
abline(coef = c(-1,1), col = "blue", lwd = 2)
abline(coef = c(1,1), col = "red", lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha("red", 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c("blue", "red"), pch = c(19), bty = "n", cex = 0.9)
@
\endminipage

@@

Moreover, all of the considered statistical tests are applied as significance tests answering 
two questions at once: (1) Should a split be performed at all? (2) If so, in which variable?
For the first question the $p$-values regarding all available covariates are compared 
to a predefined level of significance $\alpha$. Only if the smallest $p$-value is smaller
than $\alpha$ a split is performed and the covariate corresponding to this $p$-value is 
selected. Including this criterion of significance in the building process of a tree is often
referred to as \textit{pre-pruning}, however, this is not always applied in practice as 
significance is often not considered. As mentioned before, the GUIDE algorithm usually builds
large trees which are then pruned afterwards. However, for the purpose of comparing the
testing strategies rather than the whole tree algorithms in this simulation study a 
significance level is considered for all statistical tests, including the $\chi^2$-test in 
the GUIDE algorithm such that all algorithms apply pre-pruning rather than post-pruning.
The effect of these two pruning strategies are investigated in Section~\ref{sec:discussion}.

To compare the performance of the evaluated testing strategies for each step the proportion
of data sets for which the true splitting variable $Z_1$ is detected and also selected 
(for a given significance level $\alpha = 0.05$) is considered for the first scenario 
(stump).
For the second scenario (tree with 2 splits) the Adjusted Rand Index (ARI) is calculated 
as a measure of similarity between the true tree structure and the fitted model tree.

@@

\subsection{Tree}
\label{sec:tree}
To compare the performance of testing strategies in a more complex scenario they have
been evaluated on data sets based on the tree structure with two different split variables 
($Z_1$ and $Z_2$), both with the same split point $\xi$ as illustrated in Figure~\ref{fig:dgp_tree}.
